A statistical framework for whole-genome bisulfite-sequencing data

### 2015-05-03 Rmarkdown/datasets.Rmd
Fix formatting

The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA. A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (`IMR90`), with those from a pluripotent tissue, embryonic stem cells (`H1`). Each tissue was run in duplicate. While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing. In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.

@@

There are a four 'mini datasets' within the Lister data that I make some use of in my thesis. The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper. The _ADS_ sample, a human adipose tissue cell line, is the 'founder' of this mini dataset. The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line. The _ADS-adipose_ sample are "adipocytes derived from the `ADS` cells through adipogenic differentiation conditions". The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.

The second mini dataset I refer to as the _Lister-FF_ data and includes samples _FF_, _FF-iPSC\_6.9_, _FF-iPSC\_19.7_, _FF-iPSC\_19.11_ and _FF-iPSC\_11.11+BMP4_, all from the 2011 paper. The _FF_ sample, a foreskin fibroblast cell line, is the 'founder' of this this mini dataset. The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from `FF`. In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced. I believe _FF-iPSC\_6.9_ is an independently derived iPSC cell line from _FF_, although this isn't made clear in the original publication. The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (_BMP4_).

>>>>

The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA. A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (_IMR90_), with those from a pluripotent tissue, embryonic stem cells (_H1_). Each tissue was run in duplicate. While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing. In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.

@@

There are a four 'mini datasets' within the Lister data that I make some use of in my thesis. The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper. The _ADS_ sample, a human adipose tissue cell line, is the 'founder' of this mini dataset. The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line. The _ADS-adipose_ sample are "adipocytes derived from the _ADS_ cells through adipogenic differentiation conditions". The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.

The second mini dataset I refer to as the _Lister-FF_ data and includes samples _FF_, _FF-iPSC\_6.9_, _FF-iPSC\_19.7_, _FF-iPSC\_19.11_ and _FF-iPSC\_11.11+BMP4_, all from the 2011 paper. The _FF_ sample, a foreskin fibroblast cell line, is the 'founder' of this this mini dataset. The _FF-iPSC\_6.9_, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are all induced pluripotent stem cell lines derived from _FF_. In fact, _FF-iPSC\_19.7_ and _FF-iPSC\_19.11_ are subclones derived from _FF-iPSC\_19_, whose methylome was not sequenced. I believe _FF-iPSC\_6.9_ is an independently derived iPSC cell line from _FF_, although this isn't made clear in the original publication. The _FF-iPSC\_19.11+BMP4_ sample is a trophoblast cell line derived by _in vitro_ differentiating the _FF-iPSC\_19.11_ by growing a clone of it in bone morphogenic protein 4 (_BMP4_).

### 2015-05-03 Rmarkdown/methsim.Rmd
Fix formatting

Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming nature of this modelling decision. When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong. Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}). This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

>>>>

Figures \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CI} shows the overwhelming consequences of this modelling decision. When compared to the correlations of $\beta$-values from the _ADS_ input sample, we see that in all five models the correlations of $\beta$-values are far too strong. Stratifying these correlations by CpG islands, we see that the source of this problem is in regions outside of CpG islands (Figure \ref{methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}). This gives some hope that this deficiency may fixed by a more careful modelling of $\beta$-values outside CpG islands, and remains a source of ongoing work.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption[\texttt{methsim}: Correlations of $NIL = 0$ strand-collapsed $\beta$-values stratified by CpG islands]{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for \texttt{methsim} simulated data and the \emph{ADS} input sample.}
\label{fig:methsim_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Fix formatting

__TODO: Need a strong summary/conclusion. Emphasise the novel aspects of this chapter and the wide variety of datasets used.__

@@

Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA. Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.

In contrast, the somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples. The may reflect that the DNA methylome is well-established in these samples and relatively static. The increased level of partial methylation in somatic cell lines compared to somatic tissue samples may be attributable to the development of sub-clones during the culturing of the sample. This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}. Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.

>>>>

@@

Using these data, I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_). The genome-level results for the _Seisenberger_ samples are more difficult to interpret, however, since they are from pooled DNA. Nonetheless, some of the increased variability in the _Seisenberger_ data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.

In contrast, the somatic samples, particularly those from tissue samples rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples. The may reflect that the DNA methylome is well-established in these samples and relatively static. The increased level of partial methylation in somatic cell lines compared to somatic tissue samples may be attributable to the development of sub-clones during the culturing of the sample. This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}. Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.
